---
title: "Setup of Chapter 1 analyses"
output: md_document
---

```{r setup, echo = T}
library(tidyverse)
library(lme4)
```

Here I aim to simulate data in the same format as the data I will be analyzing in this chapter. There are three major types of data, summarized here.

### Simulate plant performance data

This will be a dataframe of the number of the performance of each species X site X rep in the absence of competitors. Performance will be described in two ways: 1) Raw number of seeds produced (estimated by measuring flower-heads in the field, and using size-to-seed conversions based on counting seeds in lab), and 2) Number of viable seeds produced. Note that species viability will not be at the species-by-site level, but rather at the species level. The dataframe will look something like this:

```
site  species_code  replicate num_seeds_raw num_seeds_viable
plot_740  PLER  1 144 100
plot_740  PLER  2 200 160
plot_740  PLER  3 820 700
...
```

There will data from 24 sites, 17 species, and 5 replicate individuals per species, for a total of `r 24*17*5` rows. 

Two important decisions need to be made: 1) What is to be made of 0s? 2) Do seed production numbers get scaled by species? The first is important because many species simply fail to germinate or survive to seed production at many sites. The second may be important because the seed production numbers span quite a large range - from 0, to thousands. It's unclear to me whether this matters in the analysis.

In the first dataset I simulate, the seed production numbers will be drawn from a random distribution, with a different mean per species. 

```{r generate-dummy-data-random}
set.seed(1)
n_species <- 10
n_reps <- 5
n_sites <- 20
sp_codes <- LETTERS[1:n_species]
# Seed viabiliities drawn from a uniform distribution
seed_viabilities <- runif(min = 0.2, max = 0.95, n_species)
# Mean seed production of species drawn from a lognormal, plus uniform noise
mean_seedprod <- floor(rlnorm(10, 2, 4)+runif(min = 10, max = 100, 10))
# To Do:
# Make a matrix of Species By Site preference (pre-determined for this dummy analysis)
# Then, use that to scale the seed output drawn from the distribution below: 

site_preference <- matrix(NA, nrow = n_sites, ncol = n_species)
for(current_species in 1:ncol(site_preference)) {
  preference_vector <- c(sample(seq(from = 0.1, to = 1, by = 0.1)),
                         sample(seq(from = 0, to = 0.9, by = 0.1)))
  preference_vector[sample(1:n_sites, size = 5)] = 0
  if(max(preference_vector) != 1) {preference_vector[sample(1:20, size = 1)] = 1}
  site_preference[,current_species] <- preference_vector
}

seed_prod <- data.frame(site = paste0("plot_",(rep(1:n_sites, each = n_reps))),
                        replicate = rep(1:n_reps, n_sites))

for(ii in 1:n_species){
  nsp <-rpois(n = n_reps*n_sites, lambda = mean_seedprod[ii])
  nsp <- as.numeric(floor(nsp*site_preference[,ii]))
  seed_prod <- cbind(seed_prod, nsp)
  colnames(seed_prod)[ii+2] <- LETTERS[ii]
}

seed_prod_long <- tidyr::gather(seed_prod, species, seed_production, A:J) %>% tbl_df %>% 
  mutate(site = as.factor(site), replicate = as.factor(replicate))

```

Now that we have some dummy data generated, let's plot it to explore what the heck I just did

```{r plot-exploratory-data}
gg_raw_seed <- ggplot(seed_prod_long) + geom_boxplot(aes(site,seed_production+1)) + 
  facet_wrap(~species, ncol = 5) + scale_y_log10()
gg_raw_seed + theme_gray()

gg_rawseed_hist <- ggplot(seed_prod_long) + geom_histogram(aes(seed_production+1))
gg_rawseed_hist + scale_x_log10() # Note logged X-axis
```

The details are not super relevant, just that there's some distribution of species performance by site...

### Simulating environmental data

Now, we need to simulate some environmental data for each of our sites.
**For now, these variables are all random variables- but I will update them to reflect the distribution of the true environmental data. **

```{r simulate-env-data}
site_data <- data.frame(site = paste0("plot_",(1:n_sites)),
                        soil_depth_cm = round(runif(min = 5, max = 20, n = n_sites),digits = 2),
                        mg_ppm = abs(rnorm(n = n_sites, mean = 10, sd = 2)))
```

## Begin exploring models

I need to read more of the [`lme4` package vignette](https://cran.r-project.org/web/packages/lme4/vignettes/lmer.pdf) before getting too deep into this.

```{r explore-lms}
# Merge the data frame so that the environmental variables are added on.
merged_df <- left_join(seed_prod_long, site_data)
gg_seed_v_sdep <- ggplot(merged_df) + geom_point(aes(x = soil_depth_cm, y = seed_production+1)) + 
  facet_wrap(~ species, ncol = 5) + scale_y_log10()
gg_seed_v_sdep

# Model seed production with species as the random effect
lmer(data = merged_df, formula = seed_production~1|species)

# Model seed production as a function of soil depth, with species as the random effect
lmer(data = merged_df, formula = seed_production~soil_depth_cm + (soil_depth_cm|species))
```

The models above assume that the response variable (`seed_production` here) is normally distributed. We can relax this with generalized linear models, where we can allow the response to have a different family of distribution (Poisson makes sense here).

```{r generalized-linear-model}
glmer(data = merged_df, formula = seed_production~1|species, family = poisson(link = log))

glmer(data = merged_df, formula = seed_production~soil_depth_cm + (soil_depth_cm|species), family = poisson(link = log))
```
