---
title: "ML estimates of demographic parameters"
author: "Gaurav Kandlikar"
date: "February 28, 2018"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set up competition models
First we set up the models, as motived by JHRL.

### Model 1: No effect of competitors on seed production
```{r comp-model-1}
compmodel1 <- function(par) {
  mu = par[1] 
  sigma = par[2]
  pred = mu # The predicted seed output is just the mean
  llik = dnorm(lseed, log(pred), sd = sigma, log = TRUE)
  return(sum(-1*llik))
}
```

### Model 2: A global lambda, that decreases due to competition with neighbors
```{r comp-model-2}
# write a new model that has a global alpha and a global lambda
compmodel2<-function(par){
  lambda<-par[1]
  alpha<-par[2]
  sigma<-par[3]
  pred<-lambda/(1+alpha*(dens)) #this is the predictive model
  llik<-dnorm(lseed,log(pred), sd=sigma, log=TRUE) #these are the log likelihoods
  return(sum(-1*llik)) #sum of negative log likelihoods
}
```

### Model 3: Global alpha; Global lambda that gets modified by a site-specific coefficient
```{r comp-model-3}
# write a new model that has a global alpha and a global lambda that is additively modified by a site-term
compmodel3<-function(par){
  lambda <- par[1]
  alpha  <- par[2]
  sigma  <- par[3]
  site1  <- par[4]
  site2  <- par[5] 
  site3  <- par[6] 
  site4  <- par[7] 
  site5  <- par[8] 
  site6  <- par[9] 
  site7  <- par[10] 
  site8  <- par[11] 
  site9  <- par[12]
  site10 <- par[13] 
  site11 <- par[14] 
  site12 <- par[15] 
  site13 <- par[16] 
  site14 <- par[17] 
  site15 <- par[18] 
  site16 <- par[19] 
  site17 <- par[20] 
  site18 <- par[21] 
  site19 <- par[22] 
  site20 <- par[23] 
  site21 <- par[24] 
  site22 <- par[25] 
  site23 <- par[26] 
  site24 <- par[27] 
  pred<-(lambda + site1*site_matrix[, 1]+ site2*site_matrix[, 2] + site3*site_matrix[, 3] + site4*site_matrix[, 4] + site5*site_matrix[, 5] + site6*site_matrix[, 6] + site7*site_matrix[, 7] + site8*site_matrix[, 8] + site9*site_matrix[, 9] + site10*site_matrix[, 10] + site11*site_matrix[, 11] + site12*site_matrix[, 12] + site13*site_matrix[, 13] + site14*site_matrix[, 14] + site15*site_matrix[, 15] + site16*site_matrix[, 16] + site17*site_matrix[, 17] + site18*site_matrix[, 18] + site19*site_matrix[, 19] + site20*site_matrix[, 20] + site21*site_matrix[, 21] + site22*site_matrix[, 22] + site23*site_matrix[, 23] + site24*site_matrix[, 24])/(1+alpha*(dens)) #this is the predictive model
  llik<-dnorm(lseed,log(pred), sd=sigma, log=TRUE) #these are the log likelihoods
  return(sum(-1*llik)) #sum of negative log likelihoods
}
```

### Model 4: Global alpha; 24 site-specific lambda terms per species
```{r comp-model-4}
# write a new model that has a global alpha and 24 site-specific lambda terms per species
compmodel4<-function(par){
  alpha  <- par[1]
  sigma  <- par[2]
  site1  <- par[3]
  site2  <- par[4] 
  site3  <- par[5] 
  site4  <- par[6] 
  site5  <- par[7] 
  site6  <- par[8] 
  site7  <- par[9] 
  site8  <- par[10] 
  site9  <- par[11]
  site10 <- par[12] 
  site11 <- par[13] 
  site12 <- par[14] 
  site13 <- par[15] 
  site14 <- par[16] 
  site15 <- par[17] 
  site16 <- par[18] 
  site17 <- par[19] 
  site18 <- par[20] 
  site19 <- par[21] 
  site20 <- par[22] 
  site21 <- par[23] 
  site22 <- par[24] 
  site23 <- par[25] 
  site24 <- par[26] 
  pred<-(site1*site_matrix[, 1]+ site2*site_matrix[, 2] + site3*site_matrix[, 3] + site4*site_matrix[, 4] + site5*site_matrix[, 5] + site6*site_matrix[, 6] + site7*site_matrix[, 7] + site8*site_matrix[, 8] + site9*site_matrix[, 9] + site10*site_matrix[, 10] + site11*site_matrix[, 11] + site12*site_matrix[, 12] + site13*site_matrix[, 13] + site14*site_matrix[, 14] + site15*site_matrix[, 15] + site16*site_matrix[, 16] + site17*site_matrix[, 17] + site18*site_matrix[, 18] + site19*site_matrix[, 19] + site20*site_matrix[, 20] + site21*site_matrix[, 21] + site22*site_matrix[, 22] + site23*site_matrix[, 23] + site24*site_matrix[, 24])/(1+alpha*(dens)) #this is the predictive model
  llik<-dnorm(lseed,log(pred), sd=sigma, log=TRUE) #these are the log likelihoods
  return(sum(-1*llik)) #sum of negative log likelihoods
}


```


### Model 5: 24 site-specific lambda terms and 24 site-specific alpha terms per species
```{r comp-model-5}
compmodel5<-function(par){
  # The sigma is still global
  sigma  <- par[1]

  # Site specific alphas 
  alpha1  <- par[2]
  alpha2  <- par[3] 
  alpha3  <- par[4] 
  alpha4  <- par[5] 
  alpha5  <- par[6] 
  alpha6  <- par[7] 
  alpha7  <- par[8] 
  alpha8  <- par[9] 
  alpha9  <- par[10]
  alpha10 <- par[11] 
  alpha11 <- par[12] 
  alpha12 <- par[13] 
  alpha13 <- par[14] 
  alpha14 <- par[15] 
  alpha15 <- par[16] 
  alpha16 <- par[17] 
  alpha17 <- par[18] 
  alpha18 <- par[19] 
  alpha19 <- par[20] 
  alpha20 <- par[21] 
  alpha21 <- par[22] 
  alpha22 <- par[23] 
  alpha23 <- par[24] 
  alpha24 <- par[25] 
  
  # Site specific lambas 
  site1 <- par[25]
  site2 <- par[26]
  site3 <- par[27]
  site4 <- par[28]
  site5 <- par[29]
  site6 <- par[30]
  site7 <- par[31]
  site8 <- par[32]
  site9 <- par[33]
  site10 <- par[34]
  site11 <- par[35]
  site12 <- par[36]
  site13 <- par[37]
  site14 <- par[38]
  site15 <- par[39]
  site16 <- par[40]
  site17 <- par[41]
  site18 <- par[42]
  site19 <- par[43]
  site20 <- par[44]
  site21 <- par[45]
  site22 <- par[46]
  site23 <- par[47]
  site24 <- par[48]
  
  
  # Predictive model

  pred<-(site1*site_matrix[, 1]+ site2*site_matrix[, 2] + site3*site_matrix[, 3] + site4*site_matrix[, 4] + 
           site5*site_matrix[, 5] + site6*site_matrix[, 6] + site7*site_matrix[, 7] + site8*site_matrix[, 8] + 
           site9*site_matrix[, 9] + site10*site_matrix[, 10] + site11*site_matrix[, 11] + site12*site_matrix[, 12] + 
           site13*site_matrix[, 13] + site14*site_matrix[, 14] + site15*site_matrix[, 15] + site16*site_matrix[, 16] + 
           site17*site_matrix[, 17] + site18*site_matrix[, 18] + site19*site_matrix[, 19] + site20*site_matrix[, 20] + 
           site21*site_matrix[, 21] + site22*site_matrix[, 22] + site23*site_matrix[, 23] + site24*site_matrix[, 24]) /
    (1+ alpha1*compdens[, 1]+ alpha2*compdens[, 2] + alpha3*compdens[, 3] + alpha4*compdens[, 4] + 
       alpha5*compdens[, 5] + alpha6*compdens[, 6] + alpha7*compdens[, 7] + alpha8*compdens[, 8] + 
       alpha9*compdens[, 9] + alpha10*compdens[, 10] + alpha11*compdens[, 11] + alpha12*compdens[, 12] + 
       alpha13*compdens[, 13] + alpha14*compdens[, 14] + alpha15*compdens[, 15] + alpha16*compdens[, 16] + 
       alpha17*compdens[, 17] + alpha18*compdens[, 18] + alpha19*compdens[, 19] + alpha20*compdens[, 20] + 
       alpha21*compdens[, 21] + alpha22*compdens[, 22] + alpha23*compdens[, 23] + alpha24*compdens[, 24]) #this is the predictive model
  llik<-dnorm(lseed,log(pred), sd=sigma, log=TRUE) #these are the log likelihoods
  return(sum(-1*llik)) #sum of negative log likelihoods
}

```

## Read in data, munge, and run the ML models.

```{r import-data}
library(dplyr)
compdens <- read.csv("~/Dropbox/spatial_tapioca/data/performance/calculated/NCompetitors_per_plot.csv")
seedprod <- read.csv("~/Dropbox/spatial_tapioca/data/performance/seed_production_processed.csv", stringsAsFactors = F)
# replace 0s with NAs...
seedprod[which(seedprod$num_seeds_produced == 0), "num_seeds_produced"] <- NA
# maybe switch to be the minimum for the species?
seedprod <- na.omit(seedprod)
seedprod$sp_code <- as.character(seedprod$sp_code)
seedprod$sp_code[which(seedprod$sp_code=="AGSP")] <- "AGHE"
seedprod$sp_code[which(seedprod$sp_code=="CEMI")] <- "CEME"
seedprod$sp_code[which(seedprod$sp_code=="VUMI")] <- "VUMA"

compdens <- dplyr::select(compdens, -N_per_compPlot)
compdens$N_per_neighborhood <- floor(compdens$N_per_neighborhood)
seedprod <- left_join(seedprod, compdens)
# Reset the competitor density in Lambda plots to 0
seedprod[which(seedprod$plot_type == "L"), "N_per_neighborhood"] <- 0

seedprod <- seedprod %>% mutate(num_seeds_produced = as.integer(num_seeds_produced),
                    N_per_neighborhood = as.integer(N_per_neighborhood))


# Fit the four models to just AGHE right now
# Later we loop over all the species in the dataset

# Set up empty matrices to be filled during the for-loop --------------

# Model 1
mod1pars <- matrix(NA, ncol = 3, nrow = length(unique(seedprod$sp_code)))
rownames(mod1pars) <- unique(seedprod$sp_code)
colnames(mod1pars) <- c("lambda", "sigma", "convergence_code")

# Model 2
mod2pars <- matrix(NA, ncol = 4, nrow = length(unique(seedprod$sp_code)))
rownames(mod2pars) <- unique(seedprod$sp_code)
colnames(mod2pars) <- c("lambda", "alpha", "sigma", "convergence_code")

# Model 3
mod3pars <- matrix(NA, ncol = 28, nrow = length(unique(seedprod$sp_code)))
rownames(mod3pars) <- unique(seedprod$sp_code)
colnames(mod3pars) <- c("lambda", "alpha", "sigma", paste0("plot_",740:763, "_effect"), "convergence_code")

# Model 4
mod4pars <- matrix(NA, ncol = 27, nrow = length(unique(seedprod$sp_code)))
rownames(mod4pars) <- unique(seedprod$sp_code)
colnames(mod4pars) <- c("alpha", "sigma", paste0("plot_",740:763, "_lambda"), "convergence_code")

# Model 5
mod5pars <- matrix(NA, ncol = 50, nrow = length(unique(seedprod$sp_code)))
rownames(mod5pars) <- unique(seedprod$sp_code)
colnames(mod5pars) <- c("sigma", paste0("plot_",740:763, "_alpha"),paste0("plot_",740:763, "_lambda"), "convergence_code")

for(sp in unique(seedprod$sp_code)) {
  # Format vectors for the current species ===========
  lseed <- seedprod %>% filter(sp_code == sp) %>% dplyr::select(num_seeds_produced) %>% log %>% unlist %>% as.numeric
  dens <- seedprod %>% filter(sp_code == sp) %>% dplyr::select(N_per_neighborhood) %>% unlist %>% as.numeric
  which_plot <- seedprod %>% filter(sp_code == sp) %>% dplyr::select(plot_num)
  which_plot <- which_plot$plot_num
  plot_yesno <- table(which_plot)
  
  # Make a site matrix- to be used by models 3, 4, and 5 =====
  site_matrix <- seedprod %>% filter(sp_code == sp) %>%
    mutate(site1 = ifelse(plot_num==740, 1, 0), 
           site2 = ifelse(plot_num==741, 1, 0),
           site3 = ifelse(plot_num==742, 1, 0),
           site4 = ifelse(plot_num==743, 1, 0),
           site5 = ifelse(plot_num==744, 1, 0),
           site6 = ifelse(plot_num==745, 1, 0),
           site7 = ifelse(plot_num==746, 1, 0),
           site8 = ifelse(plot_num==747, 1, 0),
           site9 = ifelse(plot_num==748, 1, 0),
           site10 = ifelse(plot_num==749, 1, 0),
           site11 = ifelse(plot_num==750, 1, 0),
           site12 = ifelse(plot_num==751, 1, 0),
           site13 = ifelse(plot_num==752, 1, 0),
           site14 = ifelse(plot_num==753, 1, 0),
           site15 = ifelse(plot_num==754, 1, 0),
           site16 = ifelse(plot_num==755, 1, 0),
           site17 = ifelse(plot_num==756, 1, 0),
           site18 = ifelse(plot_num==757, 1, 0),
           site19 = ifelse(plot_num==758, 1, 0),
           site20 = ifelse(plot_num==759, 1, 0),
           site21 = ifelse(plot_num==760, 1, 0),
           site22 = ifelse(plot_num==761, 1, 0),
           site23 = ifelse(plot_num==762, 1, 0),
           site24 = ifelse(plot_num==763, 1, 0)
    ) %>% dplyr::select(site1:site24) %>% as.matrix
  
  # Make a competition density matrix- to be used for Model 5 ======
  compdens <- seedprod %>% filter(sp_code == sp) %>%
  mutate(site1 = ifelse(plot_num==740, N_per_neighborhood, 0), 
         site2 = ifelse(plot_num==741, N_per_neighborhood, 0),
         site3 = ifelse(plot_num==742, N_per_neighborhood, 0),
         site4 = ifelse(plot_num==743, N_per_neighborhood, 0),
         site5 = ifelse(plot_num==744, N_per_neighborhood, 0),
         site6 = ifelse(plot_num==745, N_per_neighborhood, 0),
         site7 = ifelse(plot_num==746, N_per_neighborhood, 0),
         site8 = ifelse(plot_num==747, N_per_neighborhood, 0),
         site9 = ifelse(plot_num==748, N_per_neighborhood, 0),
         site10 = ifelse(plot_num==749, N_per_neighborhood, 0),
         site11 = ifelse(plot_num==750, N_per_neighborhood, 0),
         site12 = ifelse(plot_num==751, N_per_neighborhood, 0),
         site13 = ifelse(plot_num==752, N_per_neighborhood, 0),
         site14 = ifelse(plot_num==753, N_per_neighborhood, 0),
         site15 = ifelse(plot_num==754, N_per_neighborhood, 0),
         site16 = ifelse(plot_num==755, N_per_neighborhood, 0),
         site17 = ifelse(plot_num==756, N_per_neighborhood, 0),
         site18 = ifelse(plot_num==757, N_per_neighborhood, 0),
         site19 = ifelse(plot_num==758, N_per_neighborhood, 0),
         site20 = ifelse(plot_num==759, N_per_neighborhood, 0),
         site21 = ifelse(plot_num==760, N_per_neighborhood, 0),
         site22 = ifelse(plot_num==761, N_per_neighborhood, 0),
         site23 = ifelse(plot_num==762, N_per_neighborhood, 0),
         site24 = ifelse(plot_num==763, N_per_neighborhood, 0)
  ) %>% dplyr::select(site1:site24) %>% as.matrix
  
  # If there's fewer than 4 measured plants at a site, set it to zero later.
  set_to_na <- as.vector(which(colSums(site_matrix) < 4))
  
  # The model runs start here --------------
  
  # Model 1 ======
  # Set the mean and SD of lseed as the starting points for the first optimization
  par = c(mean(lseed), sd(lseed))
  # cat("Mean and SD for species:\n"); par
  for(i in 1:25){
    testcomp1<-optim(par,compmodel1)
    par<-testcomp1$par
    if(testcomp1$convergence==0){break}
  }
  print(paste(sp, ": Model 1 converged on rep", i))
  mod1pars[sp,] <- c(testcomp1$par, testcomp1$convergence)

  # Model 2 ========
  # Set mean and SD from model 1, and alpha as 0.001 to run model 2
  # Model 2 expects parameters in the order c(lambda, alpha, sigma)
  par<-c(testcomp1$par[1],0.001,testcomp1$par[2])
  for(i in 1:25){
    testcomp2<-optim(par, compmodel2, lower = c(1, 0.0000001, 0.00000001), method="L-BFGS-B")
    par<-testcomp2$par
    if(testcomp2$convergence==0){break}
  }
  print(paste(sp, ": Model 2 converged on rep", i))
  mod2pars[sp,] <- c(testcomp2$par, testcomp2$convergence)

  # Model 3 ========
  # Set the estimated global lambda, sigma, and global alpha as the beginning pars for this model; 
  # The site-specific modifiers of lambda are initialized at 0.01
  # Model 3 expects parameters in the order c(lambda,alpha, sigma, site effects*24)
  par <- c(testcomp2$par[1],testcomp2$par[2], testcomp2$par[3], rep(0.01, 24))
  for(i in 1:50){
    testcomp3<-optim(par, compmodel3, method="L-BFGS-B", lower=c(1,0,0.000001, rep(0, 24)),
                     control=list(maxit=10000), hessian=TRUE)
    par<-testcomp3$par
    if(testcomp3$convergence==0){break}
  }
  print(paste(sp, ": Model 3 converged on rep", i))
  # testcomp3$par[set_to_na+3] <- 0
  pars_to_save <- testcomp3$par
  pars_to_save[set_to_na+3] <- NA
  mod3pars[sp,] <- c(pars_to_save, testcomp3$convergence)
  
  # Model 4 =======
  # Set the estimated global lambda as the initial value of the site specific lambdas;
  # The sigma and alphas as estimated in model 2 as well.
  # Model 4 expects parameters in the order c(alpha, sigma, lambdas*24)
  par <- c(testcomp2$par[2], testcomp2$par[3], rep(testcomp2$par[1], 24))
  for(i in 1:50){
    testcomp4<-optim(par,compmodel4,  method="L-BFGS-B", lower=c(0,0.0000000001, c(rep(1,24))),
                     control=list(maxit=10000), hessian=TRUE)
    par<-testcomp4$par
    if(testcomp4$convergence==0){break}
  }
  print(paste(sp, ": Model 4 converged on rep", i))
  pars_to_save <- testcomp4$par
  pars_to_save[set_to_na+2] <- NA

  mod4pars[sp,] <- c(pars_to_save, testcomp4$convergence)
  
  # Model 5 =========
  # Set the parameters based on the output from Model 4- sigma is mod4[2], testcomp[1] is the alpha est, 
  # and 3:26 are the site-specific lambda estimates
  # Model 5 expects parameters in the order c(sigma, alphas*24, lambdas*24)
  par <- c(testcomp4$par[2], rep(testcomp4$par[1], 24), testcomp4$par[3:26])
  for(i in 1:200){
    testcomp5<-optim(par,compmodel5,  method="L-BFGS-B", lower=c(0.0000001, rep(0.0000000001, 24), rep(1, 24)),
                     control=list(maxit=10000), hessian=TRUE)
    par<-testcomp5$par
    if(testcomp5$convergence==0){break}
  }
  print(paste(sp, ": Model 5 converged on rep", i))
  pars_to_save <- testcomp5$par
  pars_to_save[set_to_na+1] <- NA
  pars_to_save[set_to_na+25] <- NA

  mod5pars[sp,] <- c(pars_to_save, testcomp5$convergence)
  
  # Plot the seed production
  # Make a vector of the predicted seed output- this is kind of tricky given the spatial goings on
  # Predicted seed output should be (Lambda_for_plot)/(1+alpha_for_plot*plot_density)
  # but we need things to repeat the right number of times...
  
  lambdas_per_site <- rep(mod5pars[sp, 26:49], colSums(site_matrix))
  alphas_per_site <- rep(mod5pars[sp, 2:25], colSums(site_matrix))
  
  pred_lseed <- c(as.matrix(lambdas_per_site*site_matrix)/
                    (1+alphas_per_site*compdens))[c(as.matrix(lambdas_per_site*site_matrix)/(1+alphas_per_site*compdens)) > 0]
  plot(x = lseed, y =pred_lseed)
  }

# Save the outputs ===========
# NOTE! Double check these outputs. Why are the convergence codes 1 now???
# write.csv(mod1pars, "model_outputs/mod1pars.csv")
# write.csv(mod2pars, "model_outputs/mod2pars.csv")
# write.csv(mod3pars, "model_outputs/mod3pars.csv")
# write.csv(mod4pars, "model_outputs/mod4pars.csv")
# write.csv(mod5pars, "model_outputs/mod5pars.csv")
```

## To Do

Here's some things that are yet to be done:   

 - [ ] Figure out exactly how to use the Hessian that is produced from testcomps3,4,5. I have a vague understanding that the square root of the diagonals of the inverse of the matrix represent the standard error of the ML estimate.   
 - [ ] Add code to plot the observed seed output vs. the seed output expected from the models  
 - [ ] Compare the output from these models with a second analysis, of estimating all of these parameters from a call to `lmer` with `species*site*competitor density` as the predictor.
 - [x] Up the max replicates of model 5 from 100 to 200, to see if the few remaining species that don't converge eventually do (maybe this doesn't matter in the end, can't hurt to try.)  