---
title: "Doing the analyses on real data"
author: "Gaurav Kandlikar"
date: "February 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Here I begin to recreate the dummy analyses presented in [this document](dummy-dataset-analysis.md)

```{r}
library(tidyverse); theme_set(theme_bw())
library(lme4)
library(vegan)
library(glmmTMB)
library(bbmle)
```

Let's import in the performance data and clean it up to our needs.

```{r manage-performance-data}

dat <- read_delim("~/Dropbox/spatial_tapioca/data/performance/seed_production_processed.csv", 
                  col_names = T, delim = ",")

dat <- dat %>% filter(plot_type == "L") %>% 
  mutate(plot_num = paste0("plot_", plot_num),
               replicate = as.factor(replicate),
               num_seeds_produced = ifelse(is.na(num_seeds_produced), 0, num_seeds_produced)) %>%
  rename(species = sp_code, seed_production = num_seeds_produced, site = plot_num) %>% select(-plot_type)

dat

gg_raw_seed <- ggplot(dat) + geom_boxplot(aes(site,seed_production+1)) + 
  facet_wrap(~species, ncol = 5) + scale_y_log10()
gg_raw_seed

gg_rawseed_hist <- ggplot(dat) + geom_histogram(aes(seed_production+1))
gg_rawseed_hist + scale_x_log10() # Note logged X-axis
```

Now we import the environmental data

```{r}
env_dat <- read_delim("~/Dropbox/spatial_tapioca/data/environmental/all_environmental_data.csv", 
                      delim = ",") 
env_dat <- env_dat %>% rename(site = plot) %>% mutate(site = paste0("plot_", site))
# skimr::skim(env_dat) %>% select(-missing, -complete, -n)

plot(env_dat %>% select(-site, -lat, -lon, -type, - microsite, -ele), 
     pch = 21, bg = alpha("black", .25))
```


Before going too far into the models, let's do an NMDS- there's a lot of soil variables to consider, so it might be wise to reduce dimensionality if there are colinear columns.

```{r soil-nmds}
# Subset the dataset to filter out some of the columns that are factors, or are repetitive...
env_dat_c <- env_dat %>% select(-type, -microsite, -lat, -lon, -Tmin, -ele) %>% 
  as.data.frame %>% tibble::column_to_rownames("site") 
soilmds <- metaMDS(env_dat_c)
plot(soilmds)
orditorp(soilmds, display="species", cex = 1.2)
plot(soilmds)
orditorp(soilmds, display="sites")

# Merge in the MDS scores into the env_dat dataframe
env_dat <- left_join(env_dat, soilmds$points %>% as.data.frame %>% tibble::rownames_to_column("site"))

# Take a look at what the distribution looks like. 
# skimr::skim(env_dat %>% select(MDS1, MDS2))
# Merge in the environmental data with the performance-by-site data, for easier modeling later. 
merged_df <- left_join(dat, env_dat)
```

Let's do some exploratory plotting of seed production vs. environmental factors
```{r plot-seedprod-v-env, eval = F}
gg_seed <- ggplot(merged_df, aes(y = seed_production+1)) + 
  facet_wrap(~ species, ncol = 5) + scale_y_log10()

gg_seed + geom_point(aes(x = depth))
gg_seed + geom_point(aes(x = Tmax))
gg_seed + geom_point(aes(x = organic_matter_ENR))
gg_seed + geom_point(aes(x = pH))
gg_seed + geom_point(aes(x = CEC_meq_100g))
gg_seed + geom_point(aes(x = soil_moisture))
gg_seed + geom_point(aes(x = Nitrate_ppm))
gg_seed + geom_point(aes(x = Mg_ppm))
gg_seed + geom_point(aes(x = Ca_ppm))
gg_seed + geom_point(aes(x = sand))
gg_seed + geom_point(aes(x = NH4_N_ppm))
gg_seed + geom_point(aes(x = MDS1))
gg_seed + geom_point(aes(x = MDS2))

```


And let's do a first pass on modeling seed production as a function of each environmental variable independently.

```{r bunch-of-glmers, eval = F}
glmer(data = merged_df, formula = seed_production~depth + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~Tmax + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~organic_matter_ENR + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~pH + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~CEC_meq_100g + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~soil_moisture + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~Nitrate_ppm + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~Mg_ppm + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~Ca_ppm + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~sand + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~NH4_N_ppm + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~MDS1 + (1|species), family = poisson(link = log))
glmer(data = merged_df, formula = seed_production~MDS2 + (1|species), family = poisson(link = log))
```

An alternative (and I'm not sure at this stage which is the better approach to this problem) is to do a complete model.

```{r complete-glmer-model, eval = F}
glmer(data = merged_df, formula = seed_production~depth+Tmax+organic_matter_ENR+pH+CEC_meq_100g+soil_moisture+
        Nitrate_ppm + Mg_ppm + Ca_ppm + sand + NH4_N_ppm + MDS1 + MDS2 + (1|site), family = poisson)
```


## I think I need to take a step back.
There's a ton of complexity in these models and I'm getting a little overwhelmed by the number of options. So I want to take a step back and break down what I am trying to do here. 

The core question for this portion of the analysis is **whether seed production responds to site-variation, and whether species differ in their responses**. Let's break things down and focus on this question, without worrying about environmental variation for now.

I will go back to the data frame `dat`, which has the following structure: 

```{r show-dat}
dat %>% head
```

Remember from above that `seed_production` is count data and, based on the histogram, seems to have a zero-inflation problem. I will use the modeling functionality in the package **`glmmTMB`** to tackle this dataset. 

```{r, eval = F}
# First, a poisson model with species and plot as fixed effect, replicate as random effect  

# The zero-eth model just has a species effect and no site effect
pm0 = glmmTMB(seed_production~species + (1|replicate), dat, family=poisson)
pm1 = glmmTMB(seed_production~species + site + (1|replicate), dat, family=poisson)
# pm2 = glmmTMB(seed_production~species * site + (1|replicate), dat, family=poisson)


nb0 = glmmTMB(seed_production~species + (1|replicate), dat, family=nbinom2)
nb1 = glmmTMB(seed_production~species + site + (1|replicate), dat, family=nbinom2)
nb2 = glmmTMB(seed_production~species * site + (1|replicate), dat, family=nbinom2)
zipm0 = glmmTMB(seed_production~species +(1|replicate), zi=~species, dat, family=poisson)
# zipm1 = glmmTMB(seed_production~species + site +(1|replicate), zi=~species, dat, family=poisson)
zipm1 = glmmTMB(seed_production~species + site +(1|replicate), zi=~species, dat, family=nbinom2)
zipm2 = glmmTMB(seed_production~species + site +(1|replicate), zi=~species + site, dat, family=nbinom2)
# zipm3 = glmmTMB(seed_production~species +(1|replicate), zi=~species, dat, family=poisson)


bbmle::AICtab(pm0, pm1, 
              nb0,nb1,nb2,
              zipm0, zipm1,zipm2)
```
